\titletext{}

### 2021-02-01 manuscript/regional_befr_manuscript.tex
Made TRY exploration production ready

% BEFR is important to study, setting the scene - DONE
Understanding the relationship between biodiversity and ecosystem function has become a central endeavour in ecological science, as we seek to predict the consequences of global biodiversity change \citep{}. Over the past two decades, study of the Biodiversity-Ecosystem Function Relationship (BEFR) has grown from small-scale experimental studies \citep{Cardinale2009, Tilman1994, Tilman2014}, to observational studies in natural ecosystems. While positive BEFRs which align with theory have been frequently reported \citep{}, as research has expanded a complex picture has emerged whereby the strength and direction of the BEFR varies depending on the ecosystem studied \citep{}, the ecosystem function(s) of interest \citep{Hector2007}, and the inclusion of environmental covariates in statistical models \citep{Vila2005}. The goal now should be to study the BEFR in different environmental and ecological contexts, in order to develop an ecosystem-agnostic understanding of the complex interactions between biodiversity, abiotic environment, and ecosystem function.

@@

There has been extended debate on whether positive BEFRs should be expected in all environmental and ecological contexts \citep{}. In temperate forests, wet tropical forests and grasslands, where the majority of BEFR studies in natural systems have been conducted \citep{}, the mechanism of niche complementarity, which contributes the majority of the observed biodiversity effect \citep{Poorter2015, Sande2017a, Wright2017}, hinges on the condition that intraspecific competition is the limiting factor to ecosystem functioning \citep{}, but this may not hold true in all systems. In mesic savannas for instance, disturbance by fire \citep{Lehmann2014} and herbivory \citep{Sankaran2008, Levick2009} reduces woody stem density and biomass, diminishing competitive interactions between individuals, allowing competitors to co-exist where they would normally be excluded \citep{Grime1979, Grace1990}. Instead, stress tolerance and the functional contribution of particular species (selection effects) may be the predominant forces influencing ecosystem function in these ecosystems \citep{Lasky2014, Tobner2016}. It is possible that a threshold woody stem density exists below which the lack of competition precludes the detection of a biodiversity effect on ecosystem function. Additionally, facilitation effects may play a larger role in driving the BEFR in stressful environments like mesic savannas. \citet{Ratcliffe2017} found stronger positive relationships between tree species richness and various ecosystem functions with increasing aridity. They demonstrate that competition diminishes where environmental stress limits species' abundances. This potential mismatch in the contribution of different mechanisms to the BEFR among resource-, disturbance-, and competition-limited ecosystems requires further investigation if we are to derive a generalisable BEFR.

Disturbance-driven mesic savannas and open canopy woodlands cover >20\% of the global land surface \citep{Pennington2018}, and represent the dominant vegetation type in southern Africa, spanning >4 million km\textsuperscript{2} \citep{White1987, Ratnam2011, Ryan2016} (\autoref{clust_map}). Despite their global importance, African mesic savannas are severely under-represented in BEFR studies \citep{Clarke2018}. A number of studies, all of which were restricted in spatial scope to a small region of miombo woodland, found that above-ground woody carbon/biomass stocks correlate positively with tree species richness \citep{McNicol2018, Shirima2015, Mutowo2012}. These studies however, lacked the breadth of sampling required to account for the interacting effects of variation in abiotic environment, disturbance regime, and biogeography which are expected to affect the BEFR.

% Environment affects the BEFR, so must be accounted for.
Southern African woody savannas occur over a wide range of precipitation, temperature, soil and disturbance conditions \citep{Frost1996}. Environmental heterogeneity has been shown to affect both woody biomass and tree species diversity independently, across a number of different biomes \citep{Spasojevic2014, Michaletz2018, Michaletz2014}. It is important therefore to account for environmental heterogeneity and understand how it influences both biomass and biodiversity to effectively model and correctly attribute the effects of biodiversity on woody biomass. \citet{Sankaran2005} and \citet{Lehmann2014} both report that total precipitation sets the upper limit for woody biomass in African savannas. \citet{Lehmann2014} also report complex indirect relationships between climate, disturbance by fire and woody biomass, demonstrating the need for directional multi-facetted modelling techniques to properly account for the effects of climate.

% In this study :::
In this study, we make the first known estimation of the Biodiversity-Ecosystem Function Relationship (BEFR) across a sub-continental area of disturbance-driven mesic woody savannas in southern Africa. We aim to understand the synergistic effects of environmental and biotic drivers of variation in the BEFR in this system, with a view to contributing to a general model of the BEFR in wooded ecosystems. We posit three hypotheses: (1) water availability and soil fertility will indirectly positively affect woody biomass via an increase in tree species diversity, (2) the effect of tree species diversity on woody biomass will increase with stem density (number of stems ha\textsuperscript{-1}), due to increased niche complementarity with stem density. In addition, we expect that an increase in disturbance by fire will decrease stem density and therefore competition, weakening the effect of tree species diversity on woody biomass. Finally, we expect that (3) tree species diversity will increase tree structural diversity (i.e. physiognomic diversity), providing an indirect path by which tree diversity increases woody biomass.

@@

% Situation - DONE
The study used \nplots{} woodland monitoring plots from a larger pool of 5395 plots in the larger SEOSAW database \citep{seosaw_web}, located across 10 countries within southern Africa in the miombo ecoregion (\autoref{clust_map}, \citealp{White1987}). The study area spans the core climate space of the region, with a precipitation gradient from c. 460 mm y\textsuperscript{-1} in southern Mozambique and southern Zimbabwe to c. 1700 mm y\textsuperscript{-1} in northern Zambia, Malawi and northern Mozambique. A 2D convex hull of Mean Annual Precipitation (MAP) and Mean Annual Temperature (MAT) of the study sites covers \hullcover{}\% of the pixel-wise climate space of the miombo woodland ecoregion \citep{White1987}, using WorldClim estimates of Mean Annual Temperature (MAT, BIO1) and Mean Annual Precipitation (MAP, BIO12) between 1970 and 2000 with a pixel size of 30 arc seconds (926 m at equator) \citep{Fick2017}.

@@

To explore the hypothesis that niche complementarity effects increase in strength as stem density increases, we repeatedly sub-sampled the available plot dataset to create \subn{} datasets of similar size with varying median stem density. We used each of these datasets separately to fit the model including only tree species and structural diversity latent variables to predict AGB. We excluded the effect of stem density on AGB and the correlation between stem density and species diversity from this model as we deliberately controlled stem density in our subsampling. We then examined how the unstandardised path coefficients for each path in the SEM varied according to the median stem density of subsampled datasets. Preliminary analyses that included herbivore biomass \citep{Hempson2017} did not converge, possibly due to the spatially coarse nature of the available data, we therefore did not include herbivory in our final model. We incorporated environmental covariates into our model to understand the relative effects of water availability, soil fertility and disturbance on AGB both directly and indirectly via species diversity and stem density. We compared standardised path coefficients between paths in the model to understand the relative contribution of each path to explain variance in AGB. Vegetation type specific models could not be reliably fitted for this more complex model specification with environmental covariates, due to sample size issues and because some vegetation types were narrow in their climate space, leading to a lack of environmental variation, particularly in the water availability latent variable.

@@

% Correlations
Pairwise correlations between all observed variables used in the Structural Equation Models (SEMs) showed that all tree species diversity and structural diversity variables had moderate positive correlations with AGB (\autoref{corr_mat}, Fig. S3). Stem density had the strongest correlation with AGB of all variables considered (\ccib{}). Environmental variables had weaker correlations with AGB than diversity variables, with all environmental variables having significant correlations with AGB, except fire frequency. The direction of these correlations was used as a test of our assumptions for the direction of influence of latent variables later used in the SEMs. MAP had positive correlations with all tree species diversity and structural diversity variables. Tree species diversity variables had clear positive correlations with stem density (species richness: \ccsi{}; Shannon equitability: \ccei{}), but structural diversity variables showed weak correlations with stem density (DBH CoV: \ccdvi{}, Height CoV: \cchvi{}).

@@

In our sub-sampling of the plot dataset by stem density, we found an increasing positive effect of tree species diversity on AGB as stem density increased (\autoref{sem_struc_stems_ha}). There appears to be a minimum stem density threshold at c. 180 trees ha\textsuperscript{-1} below which there appears to be a reasonably constant baseline effect of tree diversity on biomass. The effect of structural diversity on AGB appears to remain constant with increasing stem density. The indirect effect of tree species diversity on AGB via structural diversity increases as stem density increases.

>>>>

% BEFR is important to study, setting the scene - DONE
Understanding the relationship between biodiversity and ecosystem function has become a central endeavour in ecological science, as we seek to predict the consequences of global biodiversity change \citep{}. Over the past two decades, study of the Biodiversity-Ecosystem Function Relationship (BEFR) has grown from small-scale experimental studies mostly in temperate grasslands \citep{Cardinale2009, Tilman1994, Tilman2014}, to observational studies in natural ecosystems \citep{Plas2019}. While positive BEFRs which align with theory have been frequently reported \citep{}, as research has expanded a complex picture has emerged whereby the strength and direction of the BEFR varies depending on the ecosystem studied \citep{}, the ecosystem function(s) of interest \citep{Hector2007}, and the inclusion of environmental covariates in statistical models \citep{Vila2005}. The goal now should be to study the BEFR in different environmental and ecological contexts, in order to develop an ecosystem-agnostic understanding of the complex interactions between biodiversity, abiotic environment, and ecosystem function.

@@

There has been extended debate on whether positive BEFRs should be expected in all forest systems \citep{}. In temperate and wet tropical forests, where the majority of BEFR studies in natural forest ecosystems have been conducted \citep{Plas2019}, the mechanism of niche complementarity, which contributes the majority of the observed biodiversity effect \citep{Poorter2015, Sande2017a, Wright2017}, hinges on the condition that intraspecific competition between trees is the limiting factor to ecosystem functioning \citep{}, but this may not hold true in all systems.

Disturbance-driven mesic savannas and open canopy woodlands cover >20\% of the global land surface \citep{Pennington2018, Solbrig1996}, and represent the dominant vegetation type in Africa, spanning >4 million km\textsuperscript{2} \citep{White1987, Ratnam2011, Ryan2016} (\autoref{clust_map}). Taken together, the above- and below-ground carbon stored in African mesic savannas is comparable to that found in the wet forests of the Congo basin \citep{Houghton2009, Mayaux2008, Spawn2020}. Despite their global importance however, African mesic savannas are severely under-represented in BEFR studies \citep{Clarke2018}. In mesic savannas disturbance by fire \citep{Lehmann2014} and herbivory \citep{Sankaran2008, Levick2009} reduces woody stem density and biomass, diminishing competitive interactions between individuals, allowing competitors to co-exist where they would normally be excluded \citep{Grime1979, Grace1990}. Instead, stress tolerance and the functional contribution of particular species (selection effects) may be the predominant biotic forces influencing ecosystem function in these ecosystems \citep{Lasky2014, Tobner2016}. It is possible that a threshold woody stem density exists below which the lack of competition precludes the detection of a biodiversity effect on ecosystem function.  Additionally, facilitation effects may play a larger role in driving the BEFR in stressful environments like mesic savannas. \citet{Ratcliffe2017} found stronger positive relationships between tree species richness and various ecosystem functions with increasing aridity. They demonstrate that competition diminishes where environmental stress limits species' abundances. This potential mismatch in the contribution of different mechanisms to the BEFR among resource-, disturbance-, and competition-limited ecosystems requires further investigation if we are to derive a generalisable BEFR.

A number of studies, all of which were restricted in spatial scope to a small region of miombo woodland, the dominant mesic savanna type in southern Africa \citep{}, found that above-ground woody carbon/biomass stocks correlate positively with tree species richness \citep{McNicol2018, Shirima2015, Mutowo2012}. These studies however, lacked the spatial extent required to account for the interacting effects of variation in abiotic environment, disturbance regime, and biogeography which are expected to affect the BEFR. Studies of the BEFR often find that within a small spatial extent (<50 km), biodiversity shows a strong effect on ecosystem function, but at broader extents (>1000s km) biodiversity effects pale in significance compared to abiotic factors such as climate \citep{Gonzalez2020}. In West Africa, \citet{Mensah2020} found that woodlands and some forests showed a positive effect of tree species richness on above-ground carbon, while open savannas did not, implying that stem density and the presence of dominant large trees may affect the strength of the observed BEFR. In the Brazilian Cerrado savanna, \citet{Loiola2015} found that disturbance by fire reduced tree productivity both directly and indirectly via its effect on functional trait values that are associated with limited productivity. In contrast however, \citet{Carvalho2014} also working in the Brazilian Cerrado, found that while disturbance by fire did reduce soil fertility, which is often closely related to productivity in savannas, functional trait diversity was unaffected by fire frequency. Other studies focussing on the herbaceous diversity in disturbance-prone savannas and grasslands in North America and Europe have shown that disturbance by fire, mowing and herbivory reduces herbaceous productivity \citep{Grace2007}, and allows for weak competitors to co-exist \citep{Mason2011}. While these studies together offer some glimpse into how the BEFR may operate in disturbance-prone systems, none provide a full and comprehensive assessment of the interacting effects of diversity, abiotic environment, disturbance and ecosystem function. Additionally, due to differences in community assembly, evolutionary history and contemporary drivers between the neotropics and the African tropics, inferences from one continent cannot necessarily be applied to the other \citep{Dexter2015}.

% Environment affects the BEFR, so must be accounted for.
As well as variation in disturbance regime, southern African savannas occur over a wide range of precipitation, temperature, and soil conditions \citep{Frost1996}. Environmental heterogeneity has been shown to affect both woody biomass and tree species diversity independently, across a number of different biomes \citep{Spasojevic2014, Michaletz2018, Michaletz2014}. It is important therefore to account for environmental heterogeneity and understand how it influences both biomass and biodiversity to effectively model and correctly attribute the effects of biodiversity on woody biomass. \citet{Sankaran2005} and \citet{Lehmann2014} both report that total precipitation sets the upper limit for woody biomass in African savannas. \citet{Lehmann2014} also report complex indirect relationships between climate, disturbance by fire and woody biomass, demonstrating the need for directional multi-facetted modelling techniques to properly account for the effects of climate on ecosystem function in this system.

% In this study :::
In this study, we make the first known estimation of the Biodiversity-Ecosystem Function Relationship (BEFR) across a sub-continental area of disturbance-driven mesic woody savannas in southern Africa. We aim to understand the synergistic effects of environmental and biotic drivers of variation in the BEFR in this system, with a view to generating a general model of the BEFR in disturbance-prone forested ecosystems, which is currently lacking in the BEFR literature. We posit three hypotheses: (1) water availability and soil fertility will indirectly positively affect woody biomass via an increase in tree species diversity, (2) the effect of tree species diversity on woody biomass will increase with stem density (number of stems ha\textsuperscript{-1}), as competitive interactions become more influential as stem proximity increases. In addition, we expect that an increase in disturbance by fire will decrease stem density and therefore competition, weakening the effect of tree species diversity on woody biomass. Finally, we expect that (3) tree species diversity will increase tree structural diversity (i.e. physiognomic diversity), providing an indirect path by which tree diversity increases woody biomass.

@@

% Situation - DONE
The study used \nplots{} woodland monitoring plots from a larger pool of 5395 plots in the larger SEOSAW database \citep{SEOSAW2020}, located across 10 countries within southern Africa in the miombo ecoregion (\autoref{clust_map}, \citealp{White1987}). The study area spans the core climate space of the region, with a precipitation gradient from c. 460 mm y\textsuperscript{-1} in southern Mozambique and southern Zimbabwe to c. 1700 mm y\textsuperscript{-1} in northern Zambia, Malawi and northern Mozambique. A 2D convex hull of Mean Annual Precipitation (MAP) and Mean Annual Temperature (MAT) of the study sites covers \hullcover{}\% of the pixel-wise climate space of the miombo woodland ecoregion \citep{White1987}, using WorldClim estimates of Mean Annual Temperature (MAT, BIO1) and Mean Annual Precipitation (MAP, BIO12) between 1970 and 2000 with a pixel size of 30 arc seconds (926 m at equator) \citep{Fick2017}.

@@

To explore the hypothesis that biodiversity effects on ecosystem function increase in strength as stem density increases, we repeatedly sub-sampled the available plot dataset to create \subn{} datasets of similar size with varying median stem density. We used each of these datasets separately to fit the model including only tree species and structural diversity latent variables to predict AGB. We excluded the effect of stem density on AGB and the correlation between stem density and species diversity from this model as we deliberately controlled stem density in our subsampling. We then examined how the unstandardised path coefficients for each path in the SEM varied according to the median stem density of subsampled datasets. Preliminary analyses that included herbivore biomass \citep{Hempson2017} did not converge, possibly due to the spatially coarse nature of the available data, we therefore did not include herbivory in our final model. We incorporated environmental covariates into our model to understand the relative effects of water availability, soil fertility and disturbance on AGB both directly and indirectly via species diversity and stem density. We compared standardised path coefficients between paths in the model to understand the relative contribution of each path to explain variance in AGB. Vegetation type specific models could not be reliably fitted for this more complex model specification with environmental covariates, due to sample size issues and because some vegetation types were narrow in their climate space, leading to a lack of environmental variation, particularly in the water availability latent variable.

@@

% Correlations
Pairwise correlations between all observed variables used in the Structural Equation Models (SEMs) showed that all tree species diversity (extrapolated tree species richness, Shannon equitability index) and structural diversity (coefficient of variation of DBH and height) variables had moderate positive correlations with AGB (\autoref{corr_mat}, Fig. S3). Stem density had the strongest correlation with AGB of all variables considered (\ccib{}). Environmental variables had weaker correlations with AGB than diversity variables, with all environmental variables having significant correlations with AGB, except fire frequency. The direction of these correlations was used as a test of our assumptions for the direction of influence of latent variables later used in the SEMs. MAP had positive correlations with all tree species diversity and structural diversity variables. Tree species diversity variables had clear positive correlations with stem density (species richness: \ccsi{}; Shannon equitability: \ccei{}), but structural diversity variables showed weak correlations with stem density (DBH CoV: \ccdvi{}, Height CoV: \cchvi{}).

@@

In the sub-sampling of the plot dataset by stem density, we found an increasing positive effect of tree species diversity on AGB as stem density increased (\autoref{sem_struc_stems_ha}e). There appears to be a minimum stem density threshold at c. 180 trees ha\textsuperscript{-1} below which there appears to be a reasonably constant baseline effect of tree diversity on biomass (\autoref{sem_struc_stems_ha}b). The effect of structural diversity on AGB appears to remain constant with increasing stem density (\autoref{sem_struc_stems_ha}d). The indirect effect of tree species diversity on AGB via structural diversity increases as stem density increases (\autoref{sem_struc_stems_ha}c).
